Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming

### 2022-03-08 Paper/Manuscript_Stats&Comp/reply_letter_STCO.tex
finish editing response to reviewer 1

We have simplified the section significantly but keeping references to equations in the main text of how to implement the approximation function. We have used the diagnostics tool and iterative steps to fit and perform diagnostics on the models. We have also improved the readability of the section.

@@

This case study involves an underlying function that is composed of three additive GP functions with different kernels, squared exponential and a periodic kernels. We have applied the diagnostic tool, describing the iterative steps to fit this model and performing diagnostics on the three additive GP components at the same time. This example also serves as an example of how useful our diagnostics can be in large 1D data sets.

@@

The HSGP model is definitely useful in 1D and 2D, even for very wiggly functions, unless the number of data points is very small ($n \lesssim 300$) where the regular formulation of a GP is already fast.

It is difficult to give clear guidelines of when the application of the HSGP model can result to be impractical in other multidimensional cases, basically because it ultimately depends on the wigglyness of the function to be learned which is, in most of the cases, unknown. For that reason, in fact, the use of our diagnostics tool can be essential because it allows iteratively to fit the model using minimum computational requirements in each iteration. Observing how the requirements change in each iteration can be useful for the user to be able to recognize when the approximation is unfeasible to be applied. For $D>5$ even for very smooth functions seems to be impractical.

We disagree with the reviewer that the HSGP performs poorer than the exact GP on this Leukemia data set. In fact using the diagnostic tool the user will discover easily the optimal values for $c$ and $m$ that allows for a significantly faster performance than using the exact GP. Without the guide of the diagnostic tool the user probably would had made several trials until finding out an accurate and practical solution, consuming a lot of computation requirements in this demanding 4D case.

The statement "its applicability to multidimensional cases is mostly negative" is inaccurate, as it ultimately depends on the wigglyness of the function.

@@

Solin and Sarka (2020) fundamentally focus on the mathematical development of the model, and this work is necessary to facilitate the easier use of the method for practical users. We also analyze how the key factors of the model relate each other for accurate approximations. We derive numerical functions for these relationships that allow to update the parameters, perform diagnostics, and provide clear and useful guidelines for the whole procedure. So, we make it manageable for practical users, otherwise the practical applicability of the method would be highly limited.

We expect this paper to have an impact on how widely the basis function GPs are used, as it is makes the method more usable in practice and in broad scenarios like uni-dimensional functions, multi-dimensional functions, and other problems that require repeated fitting such as cross-validation and simulation based calibration.

@@

An inaccurate approximation implies that number of basis functions $m$ is not enough and/or the boundary factor $c$ is not adequate.

We can not state anything definite about the estimated lengthscale $\hat{\ell}$ when the approximation is inaccurate, basically because it might ultimately depend on the specific frequency patterns of the signal/function to be learned, and also because $c$ must be properly set. Although, we can state that for most of the cases, if $c$ is large enough, $\hat{\ell}$ will be below that minimum supported length-scale ($\ell^*$) determined by $m$ and $c$. But, as said before, it can not be generalized to all case studies because it might depend on the particular frequency patterns of a signal. Future research will be focused on investigating potential signals with particular frequency patterns that could escape from this rule.

What we can definitely state, which is, in fact, what our diagnostics is based on, is the following:

\begin{enumerate}
\item  Observing an $\hat{\ell}$ smaller than $\ell^*$ (i.e. $\hat{\ell}$ is overestimated), and $c$ is large enough given $\ell^*$, indicates a potential problem with the approximation. This happens because in inaccurate approximation scenarios the true length-scale is smaller than $\ell^*$, so $\hat{\ell}$ will tend to be towards smaller values than $\ell^*$.

\item The other way around, that is, observing $\hat{\ell}$ larger than $\ell^*$ (i.e. $\hat{\ell}$ is underestimated), the approximation is assumed to be accurate because when the true length-scale is larger than $\ell^*$, $\hat{\ell}$ will tend to be larger than $\ell^*$. However, we do not still have full evidence that this can be generalized to all possible cases as explained above (but, we can state that this rule can be apply for most of the the practical cases). For that reason, in order to make the diagnostic fully safety we recommend re-running the model, updating parameters, until seeing predictive accuracy measures, such as RMSE, $R^2$ and ELPD, do not improve.

As largely demonstrated in the paper in Section 4.6 and Case studies, these diagnostic rules leads to a safety, useful and fast iterative procedure to fit the models with minimum computational requirements.

These arguments have been included in the paper in Section 4.5. 
\end{enumerate}

@@

For all cases showing inaccurate approximation in Fig. 5 (left), $\hat{\ell}$ is below $\ell^*$, as long as $c$ is large enough given $\ell^*$. It can be checked by computing $\ell^*$ for each case in Fig. 5 (left) by using equation (18). Following we show a few of these inaccurate cases in Fig. 5 (left), showing diagnostic $\hat{\ell} < \ell^*$ can be trusted to diagnose a bad approximation, as long as $c$ is large enough given $\ell^*$:

@@

It shows that it is essential to follow our diagnostics tool to be sure we are using valid values in each iterative step.

@@

See response in \#1.1 above.

@@

We have developed and presented a diagnostic procedure that consists in iterative steps to fit and diagnose the model. Numerical equations are used to set initial values and update the key parameters in each iterative step. The recommended diagnostic procedure is a safety and extensive procedure that allows to automatize the applicability of the method. See details in Section 4.5 and examples in Section 4.6, Case studies and Appendix C.

@@

Section 4 have been re-structured and updated by adding new subsections, such as a specific section on the theoretical evidence of near linear proportionality among the key parameters of the model, another section with the experimentation and resulting numerical diagnostic equations, another section supporting, detailing and discussing on the diagnostic rules, and another analyzing the performance of the diagnostic tool. Additionally, new subsections regarding how to proceed with other different kernels from those implemented in this study and some insights on how the method performs in multi-dimensional cases have also been added.

We have simplified and improved significantly the writing of Section 5. We have also focused the case studies more on the main goals of the paper, which are: 1) to facilitate the implementation of the method for practical users providing clear formulation of the method, and 2) and to provide a safety and general diagnostic tool to fit and diagnose the model.

@@

Diagnostic has been build for GPs with unidimensional covariance functions. Future research will focus on building analytical models that provide the diagnostics in multi-dimensional cases. However, as an approximation, we can use the diagnostic for unidimensional functions to make diagnostics individually on the different dimensions of a multidimensional GP model.

See comment \#9 above for an extended answer.

>>>>

We have simplified the section significantly but keeping references to equations in the main text of how to implement the approximation function. We have used the diagnostic tools and iterative steps to fit and perform diagnostics on the models. We have also improved the readability of the section.

@@

Following your suggestions, we improve the case study to be better integrated with the rest of the paper. This case study involves an underlying function that is composed of three additive GP functions with different kernels, two squared exponential kernels and a periodic kernel. We have now applied the diagnostic tool, describing the iterative steps to fit this model and perform diagnostics on the three additive GP components at the same time. This case study also serves as an example of how useful our diagnostics can be in large data sets with 1D HSGPs.

@@

The HSGP model is definitely useful in 1D and 2D, even for very wiggly functions, unless the number of data points is very small ($n \lesssim 300$) where the regular formulation of a GP is already reasonably fast itself.

It is difficult to give clear guidelines of when the application of the HSGP model can result to be impractical in other multidimensional cases, basically because it ultimately depends on the wigglyness of the function to be learned which is, in most of the cases, unknown. For that reason, in fact, the use of our diagnostics tool can be essential because it allows to iteratively fit the model using minimum computational requirements in each iteration. Observing how the requirements change in each iteration can be useful for the user to be able to recognize when the approximation is unfeasible to be applied. For $D>5$, even for very smooth functions, HSGPS definitely seem to be impractical. For $D<3$ it is widely useful and significantly faster than the exact GP even for very wiggly functions.

We disagree with the reviewer that the HSGP performs poorer than the exact GP on this Leukemia data set. In fact using the diagnostic tool the user will discover easily the optimal values for $c$ and $m$ that allows for a significantly faster performance than using the exact GP. Without the guide of the diagnostic tool, the user would have likely made several trials until finding an accurate and practical solution, consuming a lot of computation requirements in this highly demanding 4D case. Insofar, the understanding that "its applicability to multidimensional cases is mostly negative" is not fully accurate, as it ultimately depends on the wigglyness of the function. We now better discuss this point in various places in the paper. Specifically, in Section 4.8, we provide details on the performance of the model in multi-dimensional cases.

%the method is limited in practice to the number of dimensions and the wigglyness of the function across the different dimensions. We have observed that it can be useful for $D<5$ depending on the wigglyness of the function. .

%In this sense, the use of our recommended diagnostic allows to find the optimal values for $c$ and $m$ which allows for minimum computational requirements which can be essential on these multidimensional cases.

We have also added a new case study to Section 5 which consists of wiggly 2D and 3D simulated data sets, where have applied the diagnostic procedure too. Even though the diagnostic was built for unidimensional kernels, we demonstrate its applicability and usefulness in multidimensional cases by diagnosing the different dimensions in each iteration individually. 
%It allows us to find the optimal solution with the minimum requirements that can be essential to use the HSGP model in multidimensional cases.
In this new case study, we have also included a brief discussion and presentation of additional results concerning computational cost in 2D, 3D, and 4D.

@@

Thank you again for your thorough review. To us, also empirical analysis provides scientific insights, in particular as we can study much more complicated scenarios that we cannot feasibly approach my mathematical analysis alone. As such, we think our paper nicely complements existing research.

Solin and Sarka (2020) fundamentally focus on the mathematical development of the model, while our work is necessary to facilitate the easier use of the method for practical users. We also analyze how the key factors of the model relate each other for accurate approximations. We derive numerical functions for these relationships that allow to update the parameters, perform diagnostics, and provide clear and useful guidelines for the whole procedure. So, we make it manageable for practical users, otherwise the practical applicability of the method would be highly limited.

We expect this paper to have a strong impact on how widely basis function GPs are used. Our contribution makes these approximate GPs more practically applicable for both uni-dimensional and multi-dimensional functions, even in combination with additional challenges that require repeated fitting such as cross-validation or simulation based calibration.

@@

We have substantially extended the discussion around our proposed diagnostic. In short, an inaccurate approximation implies that the number of basis functions $m$ is not enough and/or the boundary factor $c$ is too small. Our experiments yield consistently that, if $c$ is large enough, $\hat{\ell}$ will be below that minimum supported length-scale ($\ell^*$) determined by $m$ and $c$; and observation on which our diagnostic procedure rests. We cannot rule out with certainty that and inaccurate approximation might also yield an $\hat{\ell}$ above the minimum supported length-scale ($\ell^*$), too. However, we have not encountered this case in our experiments.

As demonstrated in Section 4.6 and our case studies, our developed diagnostic rules leads to a reliable, useful, and fast iterative procedure to fit the models with minimum computational requirements. More arguments supporting this notion have been included in the paper in Section 4.5.

%We cannot state anything definitive about the estimated lengthscale $\hat{\ell}$ when the approximation is inaccurate, basically because it might ultimately depend on the specific frequency patterns of the signal/function to be learned, and also because $c$ must be properly set. Although, we can state that for most of the cases, if $c$ is large enough, $\hat{\ell}$ will be below that minimum supported length-scale ($\ell^*$) determined by $m$ and $c$. But, as said before, it cannot be generalized to all case studies because it might depend on the particular frequency patterns of a signal. Future research will be focused on investigating potential signals with particular frequency patterns that could escape from this rule.

%What we can say more definitely is the following:

%\begin{enumerate}
%\item Observing an $\hat{\ell}$ smaller than $\ell^*$, for $c$ large enough given $\ell^*$, indicates a potential problem with the approximation. This happens because in such scenarios the true length-scale is smaller than $\ell^*$, so $\hat{\ell}$ will tend to take on values smaller than $\ell^*$.

%\item The other way around, that is, observing $\hat{\ell}$ larger than $\ell^*$, the approximation is assumed to be accurate because when the true length-scale is larger than $\ell^*$, $\hat{\ell}$ will tend to be larger than $\ell^*$. However, we do not still have full evidence that this can be generalized to all possible cases as explained above (but, we can state that this rule can be apply for most of the the practical cases). For that reason, in order to make the diagnostic fully safety we recommend re-running the model, updating parameters, until seeing predictive accuracy measures, such as RMSE, $R^2$ and ELPD, do not improve.

%\end{enumerate}

@@

We agree and have made this observation and corresponding assumptions more explicit in the paper. Below, we detail some results corresponding to Figure 5 for your convenience.

For all cases showing inaccurate approximation in Fig. 5 (left), $\hat{\ell}$ is below $\ell^*$, as long as $c$ is large enough given $\ell^*$. It can be checked by computing $\ell^*$ for each case in Fig. 5 (left) by using equation (18).

%Following we show a few of these inaccurate cases in Fig. 5 (left), showing diagnostic $\hat{\ell} < \ell^*$ can be trusted to diagnose a bad approximation, as long as $c$ is large enough given $\ell^*$:

@@

Together, this demonstrates that it is essential to follow our diagnostic procedure to ensure using valid values in each iterative step.

@@

Yes. Please see our response to your comment \#1.1 above.

@@

In response, we have developed and presented a diagnostic procedure that consists of iterative steps to fit and diagnose the model. Numerical equations are used to set initial values and update the key parameters in each iterative step. The recommended diagnostic procedure allows to automatize the application of the method. In the revised manuscript, please see Section 4.5 and examples in Section 4.6, Case studies, and Appendix C, for more details.

@@

Section 4 has been re-structured and updated by adding new subsections. In particular this includes, a subsection on the theoretical evidence of near linear proportionality among the key hyperparameters of the model, a subsection with the experimentation and resulting numerical diagnostic equations, a subsection supporting, detailing, and discussing on the diagnostic rules, and a subsection analyzing the performance of the diagnostic tool. Additionally, new subsections regarding how to proceed with other different kernels from those implemented in this study and some insights on how the method performs in multi-dimensional cases have been added.

We have simplified and significantly improved the writing in Section 5. We have also focused the case studies more on the main goals of the paper, which are: 1) to facilitate the implementation of the method for practical users providing clear a formulation of the method, and 2) and to provide an easy and general diagnostic tool to achieve good approximations via HSGPs.

@@

Our diagnostic has been build for GPs with unidimensional covariance functions. Future research will focus on building analytical models that provide the diagnostic for multi-dimensional cases, as well. However, for now, we can use the diagnostic for unidimensional functions and apply it to each dimensionan of a multidimensional GP model individually, which seems to works quite well in our examples. Please see comment \#9 above for an extended answer on this topic.
